State of the Practice for Lattice Boltzmann Method Software

### 2022-01-20 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions completed for Section 1 (Introduction)

We analyze the development of Computational Fluid Dynamics (CFD) software
packages that use the Lattice Boltzmann Method (LBM). LBM packages form a family
of algorithms for simulating single-phase and multiphase fluid flows, often
incorporating additional physical complexities \citep{chen1998lattice}, such as
reflective and non-reflective boundaries. LBM considers the behavior of a
collection of particles as a single unit at the mesoscopic scale, which lies
between the nanoscopic and microscopic scales. LBM solvers predict the
positional probability of a collection of particles moving through a lattice
structure following a two step process: i) streaming, where the particles move
along the lattice via links; and, ii) colliding, where energy and momentum is
transferred among particles that collide \citep{bao2011lattice}. As an example
of the output of LBM, Figure~\ref{circularflow} presents the streamlines of
converged flow past a stationary circular cylinder with varying Reynolds
numbers. \wss{We should ask Zahra and her students for a better example.}  LBM
has several advantages over conventional CFD methods, including a simple
calculation procedure, improved parallelization, and robust handling of complex
geometries \citep{ganji2015application}.

@@

A small sample of important applications of LBM include the following: designing
fuel cells \citep{ZhangEtAl2018}, modeling groundwater flow
\citep{AnwarAndSukop2009}, and analyzing the flow of blood in the cardiovascular
system \citep{SadeghiEtAl2020}.  As these examples illustrate, LBM can be used
for tackling problems that impact such areas as environmental policy,
manufacturing, and health and safety.  Given the important applications of LBM,
users of the libraries will be concerned with software qualities like
reliability, robustness, reproducibility, performance, correctness and
verifiability.  Given the value of their time, developers that create or modify
LBM libraries will have additional quality concerns, including maintainability,
reusability, and understandability.  With the quality considerations in mind,
and the potentially overwhelming number of choices for LBM libraries, the time
is right to assess the state of the practice.  The goal of this report is to
analyze the current state of LBM software development and compare it to the
development practices employed for research software in general.  We want to
highlight the success stories that LBM developers can share amongst their
community and with the broader research software community.  We also want to
identify areas for future improvement.

To focus our efforts, we devised 11 research questions.  The questions are
inspired by the questions that underpin our data collection methodology
\citep{SmithEtAl2021}. To start with, we need to identify the relevant examples
of LBM software.  Knowing that our data collection methodology \citep{SmithEtAl2021} requires access to source code, we build that constraint into our research question:

\begin{enumerate}
	\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What LBM
	software projects exist, with the constraint that the source code must be
	available for all identified projects?
\end{enumerate}

We next wish to assess the representative software to determine how well they
apply current software development best practices.  At this point we will base
our assessment on publicly available artifacts, where artifacts are the
documents, scripts and code that we find in a project's repository. Example
artifacts include requirements, specifications, user manuals, unit test cases,
system tests, usability tests, build scripts, API (Application Programming
Interface) documentation, READMEs, license documents, process documents, and
code.  Following best practices does not guarantee popularity, so we will also
estimate which projects are most often used by the scientists.

\begin{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Which
	of the projects identified in \rqref{RQ_WhatProjects} follow current best
	practices, based on evidence found by experimenting with the software and
	searching the artifacts available in each project's repository?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
	to the most popular projects, as viewed by the scientific community?
\end{enumerate}

To understand the state of the practice we wish to learn the frequency with
which different artifacts appear, the types of development tools used and the
methodologies used for software development.  With this data, we can ask
questions about how the LBM community compares to other research software.  That
is, in what ways does LBM follow the trends from other communities, and in which
ways is it different?

\begin{enumerate}
    \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
	do LBM projects compare to research software in general with respect to
	the artifacts present in their repositories?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
	How do LBM projects compare to research software in general with
	respect to the use of tools for:
	\begin{enumerate} 
		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
	How do LBM projects compare to research software in general with respect to
	principles, processes and methodologies used?
\end{enumerate}

Only so much information can be gleaned by looking at software repositories.  To
gain additional insight, we need to talk to developers.  We need to learn their
concerns, how they deal with these concerns, and what pain points exist for
them.  We wish to know what practices are used by the top LBM projects, so that
others can potentially emulate these practices. We also wish to identify new
practices by borrowing successful ideas from other domains. For LBM developers
that may be interested in cutting edge software practices, ideas that are
proposed, but not yet in practice, should also be identified.  All of this is
covered by the questions outlined below:

\begin{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
	the pain points for developers working on LBM software projects? 
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
	do the pain points of developers from LBM compare to the pain points
	for research software in general?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
	developers:
	\begin{enumerate} 
		\item [\rqref{RQ_Concerns}.a] what software development concerns exist; and,
		\item [\rqref{RQ_Concerns}.b] what specific practices are taken to
		address these concerns?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
	What, if any, existing research software best practice, not currently used
	by LBM software, could address the concerns identified in
	\rqref{RQ_Concerns}).
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Future}:] What practices
	could be employed by future LBM developers to address the pain points found
	in \rqref{RQ_PainPoints} and the concerns identified in \rqref{RQ_Concerns}.a?

@@

We investigated the research questions by applying the general methodology
summarized in \citet{SmithEtAl2021}.  The specific application of the
methodology to LBM is reviewed in Section~\ref{methodology}, with the full
details in \citep{Michalski2021}.  The current methodology updates the approach
used in prior assessments of domains like Geographic Information Systems
\citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
Oceanographic Software \citep{smith2015state}, Seismology software
\citep{SmithEtAl2018}, and statistical software for psychology
\citep{SmithEtAl2018_StatSoft}.  \wss{add citation to medical imaging software,
if it is an option}

With our methodology, we first choose a software domain (in the current case
LBM) and identify a list of 24 software packages.  We approximately measure the
qualities of each package by filling in a grading template. Compared with our
previous methodology, the new methodology also includes repository based
metrics, such as the number of files, number of lines of code, percentage of
issues that are closed, etc.  With the quantitative data in the grading
template, we rank the software with the Analytic Hierarchy Process (AHP). After
this, as another addition to our previous methodology, we interview some of the
development teams to further understand the status of their development process.
Finally, we summarize the measurements (Section~\ref{AHPresults}), summarize the
interview answers (Section~\ref{interviewresults}), answer the research
questions and propose recommendations for LBM software development
(Section~\ref{answersquestions}).

@@

\section{Recommendations for Addressing Developer Concerns} \label{Sec_AddressConcerns}

Answers \rqref{RQ_Recommend}.  This section answers the research questions
listed in Section~\ref{secIntro} using the quantitative data presented in
Section~\ref{AHPresults}, qualitative data presented in
Section~\ref{interviewresults}, additional data from software package
repositories and artifacts, and domain expert feedback.

>>>>

We analyze the development of Computational Fluid Dynamics (CFD) software
packages that use the Lattice Boltzmann Method (LBM). LBM packages form a family
of algorithms for simulating single-phase and multiphase fluid flows, often
incorporating additional physical complexities \citep{chen1998lattice}, such as
reflective and non-reflective boundaries. LBM considers the behavior of a
collection of particles as a single unit at the mesoscopic scale, which lies
between the nanoscopic and microscopic scales. LBM solvers predict the
positional probability of a collection of particles moving through a lattice
structure following a two step process: i) streaming, where the particles move
along the lattice via links; and, ii) colliding, where energy and momentum is
transferred among particles that collide \citep{bao2011lattice}. As an example
of the output of LBM, Figure~\ref{circularflow} presents the streamlines of
converged flow past a stationary circular cylinder with varying Reynolds
numbers. \wss{Zahra can likely provide a better example.}  LBM has several
advantages over conventional CFD methods, including a simple calculation
procedure, improved parallelization, and robust handling of complex geometries
\citep{ganji2015application}.

@@

A small sample of important applications of LBM include the following: designing
fuel cells \citep{ZhangEtAl2018}, modeling groundwater flow
\citep{AnwarAndSukop2009}, and analyzing the flow of blood in the cardiovascular
system \citep{SadeghiEtAl2020}.  As these examples illustrate, LBM can be used
for tackling problems that impact such areas as environmental policy,
manufacturing, and health and safety.  Given the important applications of LBM,
users of the libraries will be concerned with software qualities like
reliability, robustness, reproducibility, performance, correctness and
verifiability.  Since their time is valuable, developers that create or modify
LBM libraries will have additional quality concerns, including maintainability,
reusability, and understandability.  With the quality considerations in mind,
and the potentially overwhelming number of choices for existing LBM libraries,
the time is right to assess the state of the practice.  The goal of this report
is to analyze the current state of LBM software development and compare it to
the development practices employed for research software in general.  We want to
highlight success stories that LBM developers can share amongst their
community, and with the broader research software community, while at the same time watching for areas for potential future improvement.

To focus our efforts, we devised 11 research questions.  The questions are used
to structure the discussion in the paper, so for each research question below we
point to the section that contains our answer.  The questions are inspired by
the research questions that underpin our data collection methodology
\citep{SmithEtAl2021}. We start with identifying the examples of LBM software
where we have access to the source code:

\begin{enumerate}
	\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What LBM
	software projects exist, with the constraint that the source code must be
	available for all identified projects? (Section~\ref{methodology})
\end{enumerate}

We next wish to assess the representative software to determine how well they
apply current software development best practices.  At this point in the
process, to remove potential user/developer bias, we will base our assessment
only on publicly available artifacts, where artifacts are the documents, scripts
and code that we find in a project's public repository. Example artifacts
include requirements, specifications, user manuals, unit test cases, system
tests, usability tests, build scripts, API (Application Programming Interface)
documentation, READMEs, license documents, process documents, and code.
Following best practices does not guarantee popularity, so we will also compare
our ranking to how the user community itself ranks the identified projects.

\begin{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Which
	of the projects identified in \rqref{RQ_WhatProjects} follow current best
	practices, based on evidence found by experimenting with the software and
	searching the artifacts available in each project's repository? (Section~\ref{AHPresults})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
	to the most popular projects, as viewed by the scientific community? (Section~\ref{repmetrics})
\end{enumerate}

To understand the state of the practice we wish to learn the frequency with
which different artifacts appear, the types of development tools used and the
methodologies used for software development.  With this data, we can ask
questions about how LBM software compares to other research software.  That is,
in what ways does LBM follow the trends from other developer communities, and in
which ways is it different?

\begin{enumerate}
    \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
	do LBM projects compare to research software in general with respect to the
	artifacts present in their repositories?
	(Section~\ref{Sec_CompareArtifacts})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
	How do LBM projects compare to research software in general with respect to
	the use of tools (Section~\ref{Sec_CompareTools}) for:
	\begin{enumerate} 
		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
	How do LBM projects compare to research software in general with respect to
	principles, processes and methodologies used? (Section~\ref{Sec_CompareMethodologies})
\end{enumerate}

Only so much information can be gleaned by looking at software repositories.  To
gain additional insight, we need to interview developers.  We need to learn
their concerns, how they deal with these concerns, and what pain points exist
for them.  We wish to know what practices are used by the top LBM projects, so
that others can potentially emulate these practices. We also wish to identify
new practices by borrowing successful ideas from other domains. For LBM
developers interested in future cutting edge practices, ideas that are proposed,
but not yet in practice, should also be identified.  The above points are
covered by the questions outlined below:

\begin{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
	the pain points for developers working on LBM software projects?
	(Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
	do the pain points of developers from LBM compare to the pain points
	for research software in general? (Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
	developers:
	\begin{enumerate} 
		\item [\rqref{RQ_Concerns}.a] what software development concerns exist; and, (Section~\ref{Sec_Concerns})
		\item [\rqref{RQ_Concerns}.b] what specific practices are taken to
		address these concerns? (Section~\ref{Sec_AddressConcerns})
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
	What, if any, existing research software best practice, not currently used
	by LBM software, could address the concerns identified in
	\rqref{RQ_Concerns}). (Section~\ref{Sec_AddressConcerns})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Future}:] What practices
	could be employed by future LBM developers to address the pain points found
	in~\rqref{RQ_PainPoints} and the concerns identified
	in~\rqref{RQ_Concerns}.a? (Section~\ref{Sec_AddressConcerns})

@@

We investigated the research questions by applying the general methodology
summarized in \citet{SmithEtAl2021}.  The specific application of the
methodology to LBM is reviewed in Section~\ref{methodology}, with the full
details in \citep{Michalski2021}.  The current methodology updates the approach
used in prior assessments of domains like Geographic Information Systems
\citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
Oceanographic Software \citep{smith2015state}, Seismology software
\citep{SmithEtAl2018}, statistical software for psychology
\citep{SmithEtAl2018_StatSoft} and medical image analysis software
\citep{Dong2021}.  \wss{add citation to medical imaging software, if it is an
option}

With our methodology we start by identifying 24 LBM software packages.  We then
approximately measure the application of best practices for each package by
filling in a grading template. Compared with our previous methodology (as used
in \citep{SmithEtAl2016} for instance), the new methodology also includes
repository based metrics, such as the number of files, number of lines of code,
percentage of issues that are closed, etc.  With the quantitative data in the
grading template, we rank the software with the Analytic Hierarchy Process
(AHP). After this, as another addition to our previous methodology, we interview
some of the development teams to further understand the status of their
development process. The quantitative and qualitative data is then used to
answer the research questions (Sections~\ref{methodology}
to~\ref{Sec_AddressConcerns}).  Finally, we summarize the threats to validity
(Section~\ref{threats}) and the final conclusions and recommendations
(Section~\ref{conclusion}).

@@

\section{Lessons from LBM Developers} \label{Sec_AddressConcerns}

Answers \rqref{RQ_Recommend}.  Recommendations for Addressing Developer
Concerns.  This section answers the research questions listed in
Section~\ref{secIntro} using the quantitative data presented in
Section~\ref{AHPresults}, qualitative data presented in
Section~\ref{interviewresults}, additional data from software package
repositories and artifacts, and domain expert feedback.
